Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-08-12 bsis_paper/paper.tex
* Rearranged sections. * Writing work on first two sections. * Cleaned up .bib file.
git-svn-id: file:///home/rblake/svn/pamg/trunk@15 c7f3fe13-8c30-0410-9127-9a9c42fe009c

@@

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
\author[NREL]{David Alber\corauthref{cor}},
\corauth[cor]{Corresponding author. Address: 1617 Cole Blvd., Mail
  Stop: 1608, Golden, CO 80401}
\ead{david\_alber@nrel.gov}
%\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
\author[UIUC]{Luke Olson}
\address[NREL]{Scientific Computing Center\\ National Renewable Energy Laboratory}
\address[UIUC]{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
\ead{lukeo@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
% \thanks[label2]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}
% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
% \address[label2]{}

\begin{abstract}
% Text of abstract
\end{abstract}

@@

% main text

\section{Introduction}
Chapters~\ref{chap:CGS} and~\ref{chap:color-based} demonstrate the
behavior of independent set-based coarse-grid selection
algorithms. One common design property among the selection algorithms
is a routine to search for vertices to become
$C$-points. Additionally, an update in the weights of vertices as the
coarse-grid selection often follows. In this chapter, new theory and
algorithms to decrease the computational cost associated with the
search and weight update procedures used in coarse-grid selection are
presented.

\section{Coarse Grid Selection}
\label{sec:coarse-grid-select}

\subsection{Problem}
\label{sec:problem}
% Outline coarse grid selection. Problem outline.

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
% Talk about independent set-based algorithms here, and give general
% independent set-based algrotihm here.

\section{CLJP-c}
Recall that CLJP-c colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
Any set of vertices in the graph of $S$ that share the same weight is
an independent set.
\end{cor}
\begin{cor}
\label{5:cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{cor}
The first corollary states that independent sets can be selected in
the graph simply by selecting sets of vertices with the same
weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{5:sec:contribution}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

\section{Coarse-Grid Selection Search and Weight Update}
\label{5:sec:s-and-w-update}
CLJP and its descendents select an independent set $D$ in each
iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
condition for selecting $D$ is that all vertices $i \in D$ must
satisfy
\begin{equation}
\label{5:eqn:D-conditions}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
\end{equation}
CLJP and CLJP-c rely on a search routine to locate locally maximal
weights and on a weight update routine to modify weights of vertices
connected to new $C$-points.

\section{Bucket Sorted Independent Set Selection}
\label{5:sec:contribution}
In this section, new techniques for searching the graph of $S$ for new
$C$-points and subsequently updating the weights of remaining vertices
are developed. The new algorithm is labeled Bucket Sorted Independent
Sets (BSIS) to reflect the data structure used.

\subsection{Coarse Grid Invariance}
\label{5:sec:cg-invariance}
Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

@@

@@

\section{Weight Update Aggregation}
\label{5:sec:agg}
Whenever a vertex weight is updated, BSIS moves the corresponding
vertex to a new location in the data structure. During the selection
of the coarse grid, the cost of the updates to the data structure is
non-trivial and, as shown in this section, unnecessary.

@@

\section{Experimental Results}
\label{5:sec:experiments}
To demonstrate BSIS, the algorithm is compared with CLJP-c. The test
problem is the 3D 7-point Laplacian on a structured grid:
\begin{eqnarray}
\label{5:eqn:Lap} -\Delta u & = & 0 \quad \textrm{on } \Omega \qquad
(\Omega = (0, 1)^3),\\
\nonumber u & = & 0 \quad \textrm{on } \partial \Omega.
\end{eqnarray}
A 7-point Laplacian is selected since it is a common initial test
problem and structured problems often lead to the largest operator
complexities for coarsening algorithms satisfying heuristic H1. By
creating larger operator complexities, the algorithms are forced to
traverse more edges in the graph, leading to more work.

@@

\section{BSIS Variants}
The ideas developed in this chapter are applicable to other coarsening
algorithms that utilize a graph coloring step, such as color-based
methods designed to satisfy heuristic H1$^\prime$. The PMIS-c1
algorithm is expected to benefit from BSIS and is expected to perform
naturally in parallel since vertex weights are not updated in PMIS.

\section{Parallelizing BSIS}
Using BSIS in a parallel algorithm presents challenges because the
parallelism in BSIS is very fine-grained. Its elegance and potential
to greatly improve the efficiency of coarse-grid selection motivates
the development of parallel algorithms incorporating BSIS. Several
alternatives are explored in this section.

@@

>>>>

\newcommand{\hp}{H1$^\prime$}

@@

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
\author[NREL]{David Alber\corauthref{cor}},
\corauth[cor]{Corresponding author. Address: 1617 Cole Blvd., Mail
  Stop: 1608, Golden, CO 80401}
\ead{david\_alber@nrel.gov}
\ead[url]{http://google.com/ie?q=david+alber+nrel\&num=1}
\author[UIUC]{Luke Olson}
\address[NREL]{Scientific Computing Center\\ National Renewable Energy Laboratory}
\address[UIUC]{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
\ead{lukeo@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
% \thanks[label2]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}
% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
% \address[label2]{}

\begin{abstract}
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The algebraic multigrid (AMG) method~\cite{Ruge1987,Brandt1982} is an
efficient numerical algorithm for finding the solution to numerous
problems, including linear systems $Ax = b$ arising from the
discretization of partial differential equations on structured and
unstructured meshes. In many cases, the computational complexity of
AMG is $O(n)$, where $n$ is the number of unknowns in the linear
system. The linear cost scaling property of AMG makes the method
attractive, especially for large problems.

AMG algorithms are generally divided into two phases: the setup phase
and the solve phase. Setup phase algorithms provide functionality for
constructing elements needed by the solve phase, such as coarse grids,
operators to transfer information between levels of the problem, and
the linear systems to be solved on each level. The solve phase applies
relaxation methods on the problems generated in the setup phase and
uses the transfer operators to move vectors between levels.

In this paper, we focus attention on coarse-grid selection algorithms
in the AMG setup phase. One class of coarse-grid selection algorithms,
independent set-based algorithms, is studied in detail. In
Section~\ref{sec:cgs}, we introduce the coarse-grid selection problems
and the goals of coarse-grid selection algorithms. The components of
independent set-based coarsening algorithms are discussed.

\section{Coarse-Grid Selection}
\label{sec:cgs}
In AMG, coarse-grid selection is the process of selecting the degrees
of freedom for the coarse level problem. In classical AMG, degrees of
freedom on a coarse level are a subset of the degrees of freedom on a
finer level. Smoothed
aggregation~\cite{Vanek1996,Vanek2001,brezina2005} forms a
coarse-level degree of freedom by aggregating several fine-level
degrees of freedom. In this paper, we study coarsening for classical
AMG.

Given a set of fine-level degrees of freedom, $\Omega_k$, coarse-grid
selection for classical AMG selects a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed for selecting the coarse grid.

Independent set-based algorithms are most numerous and all trace their
roots to the classical AMG coarsening algorithm, RS
coarsening~\cite{Ruge1987}. These methods are our focus and are
discussed in more detail in the next section.

Other methods take substantially different approaches to
coarsening. The coarsening technique in~\cite{Griebel2006} develops a
parallel coarsening algorithm that utilizes RS. On each processor, RS
selects several different coarse grids, and then one coarse grid is
selected per processor in a way to minimize special treatment on
processor boundaries. In~\cite{MacLachlan2006}, a greedy approach is
used to produce a good splitting of the unknowns into fine-grid and
coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
offer another approach for parallel coarse-grid selection by
decoupling coarse grids and alleviating the need for communication on
coarse levels. Compatible relaxation
approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
incrementally and decide when the coarse grid is sufficient through a
quality measure based on relaxation.

\subsection{Independent Set-Based Selection}
\label{sec:indep-set-based}
Independent set-based coarse grid selection algorithms rely on
\emph{strength of connection} measures to ascertain the influence
between unknowns.

Many independent set-based coarsening algorithms form $\Omega_{k+1}$
using the following heuristics.
\begin{description}
\item[H1:] For each unknown $j$ that strongly influences $F$-point
  $i$, $j$ is either a $C$-point or strongly depends on a $C$-point
  $k$ that also strongly influences $i$.
\item[H2:] The set of $C$-points needs to form a maximal independent
  set in the reduced graph of $S$ such that no $C$-point strongly
  depends on another $C$-point.
\end{description}
H1 and H2 cannot, in general, be satisfied simultaneously, so H1 is
enforced by independent set-based methods, while H2 serves as a
guideline whose purpose is to encourage selection of small
$|\Omega_{k+1}|$.

Some independent set-based coarsening algorithms use a variant of H1,
called \hp, which is a relaxed version of H1.
\begin{description}
\item[\hp] Each $F$-point must be strongly influenced by at least one
  $C$-point.
\end{description}

\section{Invariance of Selection in Independent Set-Based Methods}
\label{sec:invariance}

Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

@@

\section{Bucket Sorted Independent Set Selection}
\label{5:sec:contribution}

\subsection{CLJP-c}
Recall that CLJP-c colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
Any set of vertices in the graph of $S$ that share the same weight is
an independent set.
\end{cor}
\begin{cor}
\label{5:cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{cor}
The first corollary states that independent sets can be selected in
the graph simply by selecting sets of vertices with the same
weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{5:sec:contribution}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

\subsection{BSIS}
\label{sec:bsis}
In this section, new techniques for searching the graph of $S$ for new
$C$-points and subsequently updating the weights of remaining vertices
are developed. The new algorithm is labeled Bucket Sorted Independent
Sets (BSIS) to reflect the data structure used.

\subsection{Motivation}
\label{sec:motivation}
CLJP and its descendents select an independent set $D$ in each
iteration. Stated originally in Section~\ref{3:sec:CLJP}, the
condition for selecting $D$ is that all vertices $i \in D$ must
satisfy
\begin{equation}
\label{5:eqn:D-conditions}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i.
\end{equation}
CLJP and CLJP-c rely on a search routine to locate locally maximal
weights and on a weight update routine to modify weights of vertices
connected to new $C$-points.

@@

\subsection{Weight Update Aggregation}
\label{5:sec:agg}
Whenever a vertex weight is updated, BSIS moves the corresponding
vertex to a new location in the data structure. During the selection
of the coarse grid, the cost of the updates to the data structure is
non-trivial and, as shown in this section, unnecessary.

@@

\subsection{Experimental Results}
\label{5:sec:experiments}
To demonstrate BSIS, the algorithm is compared with CLJP-c. The test
problem is the 3D 7-point Laplacian on a structured grid:
\begin{eqnarray}
\label{5:eqn:Lap} -\Delta u & = & 0 \quad \textrm{on } \Omega \qquad
(\Omega = (0, 1)^3),\\
\nonumber u & = & 0 \quad \textrm{on } \partial \Omega.
\end{eqnarray}
A 7-point Laplacian is selected since it is a common initial test
problem and structured problems often lead to the largest operator
complexities for coarsening algorithms satisfying heuristic H1. By
creating larger operator complexities, the algorithms are forced to
traverse more edges in the graph, leading to more work.

@@

\subsection{BSIS Variants}
The ideas developed in this chapter are applicable to other coarsening
algorithms that utilize a graph coloring step, such as color-based
methods designed to satisfy heuristic H1$^\prime$. The PMIS-c1
algorithm is expected to benefit from BSIS and is expected to perform
naturally in parallel since vertex weights are not updated in PMIS.

\subsection{Parallelizing BSIS}
Using BSIS in a parallel algorithm presents challenges because the
parallelism in BSIS is very fine-grained. Its elegance and potential
to greatly improve the efficiency of coarse-grid selection motivates
the development of parallel algorithms incorporating BSIS. Several
alternatives are explored in this section.

@@

\section{Conclusions}
\label{sec:conclusions}
